---
title: "Spatial Neighbours"
---

# Tobler’s First Law of Geography

“Everything is related to everything else, but near things are more related than distant things.” — Waldo Tobler, 1970

In simple words, objects or phenomena that are closer together in space tend to be more similar or more strongly related than those far apart.

**Example:**

Temperature readings in neighboring cities are more similar than readings in cities far apart.

Houses near a park may have similar property values compared to houses far from the park.

# Why it is important

Tobler’s Law is fundamental to spatial analysis because it explains why:

1.  Spatial autocorrelation exists

Values at nearby locations are not independent.

This is a key assumption in geostatistics, spatial regression, and kriging.

2.  We can predict values at unobserved locations Using nearby observations, we can interpolate or estimate missing values.

Example: estimating rainfall at a location using nearby weather stations.

3.  Spatial patterns are meaningful

Clustering, hotspots, and diffusion patterns occur because near things are more related.

Example: disease outbreaks, crime hotspots, deforestation.

4.  It guides sampling and modeling

Knowing that nearby points are correlated helps design surveys and spatial models efficiently. Reduces redundancy and improves predictions.

# What are Spatial Neighbours?

In spatial analysis, spatial neighbours are locations or regions that are considered “close” or related in space.

The idea is that nearby locations are more likely to influence each other—this is directly related to Tobler’s First Law of Geography: “near things are more related than distant things.”

# Types of Spatial Neighbours

1.  Contiguity-based neighbours

Two regions are neighbours if they share a common boundary (or vertex).

Rook criterion: Share a common edge.

Queen criterion: Share an edge or a vertex.

Example: Two adjacent districts in a country.

2.  Distance-based neighbours

Two locations are neighbours if the distance between them is below a threshold.

Example: Weather stations within 10 km of each other.

3.  k-nearest neighbours

Each location is connected to its k closest locations, regardless of actual distance threshold.

# Why Spatial Neighbours Matter?

They capture spatial dependence in data.

They are used to create spatial weight matrices, which are central in:

Spatial regression (spatial lag, spatial error models)

Moran’s I and other spatial autocorrelation measures

Predictive modelling and kriging

# Practical Session with R

## Packages

```{r}
library(spdep)
library(sf)
```

## Data

```{r, echo=TRUE}
columbus <- st_read(system.file("shapes/columbus.gpkg", package="spData")[1], quiet=TRUE)
columbus
```

## Explore data structure

```{r, echo=TRUE}
st_crs(columbus)

```

```{r, echo=TRUE}
sf::st_geometry(columbus)

```

## Visualise data

`st_geometry()` is an `sf` function that extracts only the geometry from an sf object.

```{r, echo=TRUE}
st_geometry(columbus)
```

```{r, echo=TRUE}
plot(st_geometry(columbus))
```

\`\`\`

In-class

Calculate neighbors

## 

# Higher-order Spatial Neighbour

In-class

# Neighborhood matrix

in-class
